---
title: "Computing Variance Inflation Factor (VIF) of predictor variables"
author: "Ransmond Berchie"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Computing Variance Inflation Factor (VIF) of predictor variables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
  
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# The R VIFpkg Package



## Multicollinearity & Varaince Inflation Factor (VIF)

In Statistics, multicollinearity exists when a predictor in a multiple regression model can be linearly predicted from the other predictors with a considerable degree of accuracy. Thus, this is a phenomenon in which two or more predictor variable in a multiple regression model are highly or moderately correlated with one another.

When multicollinearity exists, it can ruin our analysis and hence limit conclusions we can draw from our results. It existence can aggravate the following outcomes in our analysis:
  
  
  * The estimated regression coefficient of any one predictor variable will depend on which other predictor variables are included in the model.

* As more predictor variables are added to the model, the precision of the estimated regression coefficients decreases.

* The predictor variables already in the model determines the marginal contribution of any one predictor variable in reducing the error sum of squares.

* Hypothesis tests for $\beta_k = 0$ may yield different conclusions depending on which predictor variables are in the model.



Multicollinearity is a major issue in observational studies, since researchers just observe values of the predictor variables without any control over them.


We can detect if multicollinearity is present in our data using these common methods:
  
  
  * The analysis exhibits the signs of multicollinearity — for instance, the estimates of coefficients vary excessively from model to model.

* The t-tests for each of the individual slopes are non-significant $(P > 0.05)$, but the overall F-test for testing all of the slopes are simultaneously $0$ is significant $(P < 0.05)$.

* The correlations among pairs of predictor variables are large.





However, it may not be enough to rely of correlations among pairs of predictors to detect multicollinearity. This is because pairwise correlations may be small, but there might be a linear dependence among three or more variables. For example, $X_1 = 3X_2 + 6X_3 + error$.

In view of this, many analysts use the \textbf{variance inflation factor (VIF)} to help detect multicollinearity.


## Variance Inflation Factor

The VIF quantifies how much the variance s inflated. When multicollinearity exists, the variance of the estimated coefficients are inflated (i.e., the precision of the estimated regression coefficients decreases). In a multiple regression model, the VIF $(VIF_j)$ for an estimated regression coefficient $(\hat{\beta_j})$ is the factor by which the variance of $\hat{\beta_j}$ is inflated due to the existence of correlation among the predictor variables in the model. 

The VIF for predictor $j$ is calculated as;

$$VIF_j = \frac{1}{1 - R^2_j}$$
  
  
  Where, $R^2_j$ is the Coefficient of Determination ($R^2$ value) for regressing predictor $j$ against all other predictor variables.

A VIF of value 1 indicates that there is no correlation among the $j^{th}$ predictor and the remaining predictor variables(i.e., the variance of predictor $j$ is not inflated). As a general rule of thumb, VIFs greater than 4 calls for investigation, and VIFs exceeding 10 indicates serious multicollinearity.


## Using the VIFsM() function to find VIF on a predictor dataset in R

### Functions

VIFsM(Data, multi = FALSE, maxim = 4)


Plot(x, maxx = 4, multi = FALSE, ...)


summarY(object, multi = FALSE)




***Arguments***

```{r, echo=FALSE}

dat.describe <- data.frame(Argument= c("Data",
"multi", "maxim", "x", "maxim", "..."), Description = c("a dataframe or matrix of predictors",
"logical stating whether to drop the variable with VIF greater than maxim, and proceed to recompute VIF until no multicollinearity exist based on maxim value defined. Default is FALSE, which just computes the VIF of the current data",
"numeric value. The maximum acceptable VIF value. Predictor variables are sequentially kicked out of the model if they have VIF greater than the maxim when multi = TRUE. Default is value 4.", "object of class VIFsM", "numeric value. The maximum acceptable VIF value. Predictor variables are sequentially kicked out of the model if they have VIF greater than the maxim when multi = TRUE. Default is value 4.", "Further arguments passed to ggplot2") )


kableExtra::kable(dat.describe)
```


## Usage of functions

### When Interested in the VIF of the current data (multi = FALSE, as in default)

```{r setup}

# Install package

# install.packages("VIFpkg_0.1.0.tar.gz", repos = NULL)

# load the VIFpkg pacakge

library(VIFpkg)

# Using Boston data from MASS package for our example

df <- data.frame(MASS::Boston[,-1])

# Find VIF values

xs <- VIFsM(df, multi=FALSE,maxim = 4)

# View the results

summarY(object=xs)


# Plot the VIF values

Plot(x = xs,maxx = 4)





```



### When Interested in checking the VIF of the data and dropping the variable with VIF > maxim

```{r, fig.height=12, fig.width=10}
# Find VIF values

xs <- VIFsM(df, multi=TRUE,maxim = 4)

# View the results

summarY(object=xs, multi = TRUE)


# Plot the VIF values

Plot(x = xs,multi = TRUE, maxx = 4)

```



***Sources***
  
  * https://online.stat.psu.edu/stat462/node/180/
  
  * https://en.wikipedia.org/wiki/Multicollinearity

